\documentclass[12pt]{article}

\usepackage{amsmath}

\input{cs486_assign_preamble.tex}

\lhead{CS 486/686}
\chead{Fall 2022}
\rhead{Academic Integrity Statement}
\cfoot{v1.0}


\hyphenation{PYTHONHASHSEED}

\begin{document}

\section*{Academic Integrity Statement}

{\color{red} Your submission on Crowdmark must include this academic integrity statement with your signature (typed or signed name). It is worth 5 marks.} 

I declare the following statements to be true:

\begin{itemize}
\item 
The work I submit here is entirely my own.

\item 	
I have not shared and will not share any of my code with anyone at any point. 

\item 
I have not posted and will not post my code on any public or private forum or website.

\item 	
I have not discussed and will not discuss the contents of this assessment with anyone at any point.

\item 
I have not posted and will not post the contents of this assessment and its solutions on any public or private forum or website. 

\item 
I will not search for assessment solutions online.

\item
I have not used an automated code assistance tool (such as GitHub Copilot).

\item 
I am aware that misconduct related to assessments can result in significant penalties, possibly including failure in the course and suspension. This is covered in Policy 71: \url{https://uwaterloo.ca/secretariat/policies-procedures-guidelines/policy-71}.
\end{itemize}

By typing or writing my full legal name below, I confirm that I have read and understood the academic integrity statement above.

\vspace{30pt}

\begin{minipage}{12cm}
\hrulefill
\end{minipage}


\end{document}